Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-10-06 paper.Rnw
results were reproducible! so I updated the text and am working on more changes to text now

% Set up for ASA data science journal
\articletype{Research Article}
\received{XX XX 2020}
\revised{XX XX XXXX}
\accepted{XX XX XXXX}
\raggedbottom

@@

First, consider the claim of localness. The weights decay relatively slowly  outside of the intersection of the 2-quantil-bin suggesting that the LIME explanation is highly influenced by points outside of the bins containing the prediction of interest. However, it is  difficult to say if the claim of localness has been violated, because a definition of a local region is not specified. Depending on what region a viewer considers to be local, an argument could be made in favor of or against a violation of localness. The explanation scatterplot raises awareness of the unclear definition of a local region with LIME. Nevertheless, it is possible to say that the weights assigned to the simulated data do not capture  the local region identified by the intersection of the quantile bins.

@@

# Train a new random forest that mimics the rtrees models 
# (or load if already trained)
bullet_rf_file = "./data/bullet-rf.rds"
if (!file.exists(bullet_rf_file)) {
  set.seed(20200923) # the day that the updated data was found :)
  bullet_rf <-
    randomForest(
      y = factor(bullet_train$samesource),
      x = bullet_train %>% select(bullet_features),
      ntree = bullet_rf_ntrees,
      mtry = bullet_rf_mtry
    )  
  saveRDS(object = bullet_rf, file = bullet_rf_file)
} else {
  bullet_rf = readRDS(file = bullet_rf_file)
}

@@

\hh{XXX minor change to emphasize that this is not a completly new model, just an expanded training set - otherwise you open yourself up to criticism of not using a validated model. The bulletxtrctr model was used in an external validation in \citet{validation}. }
\hh{Here, we train a } random forest model using the same nine comparison features as \citet{hare:2017} to predict whether  \hh{two lands come from the same source} (i.e.\ are a match) in an expanded set of \Sexpr{format(dim(bullet_train)[1], big.mark = ",")} land-to-land comparisons from \hh{two sets of} bullets in the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}. We apply the trained random forest model to 6 bullets from \hh{another set of} the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. See the supporting information for further descriptions of the data.

@@

First, focus on the explanation for case NM with 4-quantile-bins. The explanation plot from \emph{lime} shows that observed values with rough correlation (rough\_cor) greater than 0.05 and mismatches value less than or equal to 7.17 support a random forest prediction in favor of a match, and the scatter plots show that many of the simulated values with rough correlation greater than 0.05 and mismatches less than 7.17 have random forest predictions greater than 0.5. Additionally, the LIME explanation indicates that a value of cross correlation function (ccf) greater than 0.28 and less than or equal to 0.35 supports a non-match, and the scatter plots shows that the simulated values within this region have mostly random forest predictions less than 0.5.  However, the random forest assigns a probability of 0.19 to case NM, which is not explained by the LIME explanation. Instead, consider that a better explanation based on the scatterplots would be that the random forest assigned a prediction probably close to 0 since the observation has a rough correlation less than 0.5, mismatches greater than 4, and cross correlation function 0.6.

The other explanation scatterplots can be interpreted in a similar manner. Overall, the plots identify that the 4-quantile-bin explanations are not faithful to the random forest model predictions. The 4-equal-bin explanations are more faithful to the random forest, but the regions captured by the bins do not align well with the regions containing  similar probabilities produced by the random forest.

@@

The code used to produce the manuscript and the data from examples in Section \ref{application} are available in the following GitHub Repository: \href{https://github.com/goodekat/LIME-diagnostics-paper}{https://github.com/goodekat/LIME-diagnostics-paper}. \hh{\sout{ Additional information about the bullet matching data are available at XXX.}}

@@

@
\caption{Explanation scatterplots for LIME explanations using kernel density simulation for the cases M and NM of the bullet comparison.}
\label{fig:figure-B2}
\end{figure*}

>>>>

% Set up for ASA data science journal
\articletype{Research Article}
\received{XX XX XXXX}
\revised{XX XX XXXX}
\accepted{XX XX XXXX}
\raggedbottom

@@

First, consider the claim of localness. The weights decay relatively slowly outside of the intersection of the 2-quantile-bin suggesting that the LIME explanation is highly influenced by points outside of the bins containing the prediction of interest. However, it is difficult to say if the claim of localness has been violated, because a definition of a local region is not specified. Depending on what region a viewer considers to be local, an argument could be made in favor of or against a violation of localness. The explanation scatterplot raises awareness of the unclear definition of a local region with LIME. Nevertheless, it is possible to say that the weights assigned to the simulated data do not capture  the local region identified by the intersection of the quantile bins.

@@

# Train a new random forest that mimics the rtrees models 
# (or load if already trained)
bullet_rf_file = "./data/bullet-rf.rds"
if (!file.exists(bullet_rf_file)) {
  set.seed(20200923) # the day that the updated data was found :)
  bullet_rf <-
    randomForest(
      y = factor(bullet_train$samesource),
      x = bullet_train %>% select(all_of(bullet_features)),
      ntree = bullet_rf_ntrees,
      mtry = bullet_rf_mtry
    )  
  saveRDS(object = bullet_rf, file = bullet_rf_file)
} else {
  bullet_rf = readRDS(file = bullet_rf_file)
}

@@

%XXX minor change to emphasize that this is not a completely new model, just an expanded training set - otherwise you open yourself up to criticism of not using a validated model. The bulletxtrctr model was used in an external validation in \citet{vanderplas:2020b}.

\kgc{Your comment about not training a whole new model made lost of sense. Are my changes okay?}

\kge{A random forest model fit to an expanded dataset of \Sexpr{format(dim(bullet_train)[1], big.mark = ",")} land-to-land comparisons from two sets of bullets in the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009} was validated in \citet{vanderplas:2020b}. Here, we train a random forest model to mimic the one validated in model \citet{vanderplas:2020b} using the same data, model structure, and features (the nine similarity measures developed in \citet{hare:2017})}. %Here, we train a random forest model using the same \kge{data, features (the nine similarity measures developed in \citet{hare:2017}), and model structure as a random forest validated in \citet{vanderplas:2020b}} \kge{\sout{nine comparison features as \citet{hare:2017}}} to predict whether two lands come from the same source (i.e.\ are a match) in an expanded set of \Sexpr{format(dim(bullet_train)[1], big.mark = ",")} land-to-land comparisons from two sets of bullets in the James Hamby Consecutively Rifled Ruger Barrel Study \citep{hamby:2009}. 
We apply the trained random forest model to 6 bullets from another set of the Hamby study with \Sexpr{dim(bullet_test)[1]} rows of land comparisons. See the supporting information for further descriptions of the data.

@@

\kge{We \sout{First,}} focus on the explanation for case NM with 4-quantile-bins. The explanation plot from \emph{lime} shows that \kge{the} observed values \kge{of NM of \sout{with}} \kge{mismatches less than or equal to 7.17, rough correlation (rough\_cor) greater than 0.05, and D less than or equal to 0.002 support a random forest prediction in favor of a match. The scatter plots show that many of the simulated values in the rectangular regions with mismatches less than 7.17, rough correlation greater than 0.05, and distance less than or equal to 0.002 have random forest predictions greater than 0.5.} %rough correlation (rough\_cor) greater than 0.05 and mismatches value less than or equal to 7.17 support a random forest prediction in favor of a match, and the scatter plots show that many of the simulated values with rough correlation greater than 0.05 and mismatches less than 7.17 have random forest predictions greater than 0.5. Additionally, the LIME explanation indicates that a value of cross correlation function (ccf) greater than 0.28 and less than or equal to 0.35 supports a non-match, and the scatter plots shows that the simulated values within this region have mostly random forest predictions less than 0.5. 
However, the random forest assigns a probability of 0.21 to case NM, which is not explained by the LIME explanation \kge{that provides all explanations in support of a match}. Instead, consider that a better explanation based on the scatterplots would be that the random forest assigned a prediction probably \kge{below 0.5 \sout{close to 0}} since \kge{NM has observed values of mismatches less than 5, rough correlation greater than 0.25, and distance greater than 0.001.}%the observation has a rough correlation less than 0.5, mismatches greater than 4, and cross correlation function 0.6.

The other explanation scatterplots can be interpreted in a similar manner. \kge{\sout{Overall, t} T}he \kge{three plots depict situations where the explanations are faithful} %plots identify that the 4-quantile-bin explanations are not faithful to the random forest model predictions. The 4-equal-bin explanations are more faithful 
to the random forest, but the regions captured by the \kge{intersections of the} bins do not align well with the regions containing similar probabilities produced by the random forest.

@@

The code used to produce the manuscript and the data from examples in Section \ref{application} are available in the following GitHub Repository: \href{https://github.com/goodekat/LIME-diagnostics-paper}{https://github.com/goodekat/LIME-diagnostics-paper}.

@@

@
\caption{Explanation scatterplots for LIME explanations using kernel density simulation for the cases M and NM of the bullet comparison test data.}
\label{fig:figure-B2}
\end{figure*}
